pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-13 article.tex
I enjoy ERQ 1 so much better now

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

@@

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
We will exploit these functions to answer our research questions in the next sections.

@@

All \verb;pirouette;'s functions are documented. 
Useful examples and sensible defaults, 
which we will refer to in the next sections, 
can be accessed from the documentation.

@@

@@

"What is the error made by BEAST2 from a phylogeny 
using the same diversification model as it was generated by?"

We start from a tree simulated under a Yule (pure-birth) generative prior. 
We choose to use a tree with six taxa and a crown age of ten.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_yule.png}
  \caption{The Yule tree, as created by listing~\ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
  \giovanni{Remove the text in top-left corner.}
  \giovanni{I'd use a sexier topology, purely for esthetical reasons.}
\end{figure}

The first step in \verb;pirouette; is to simulate a DNA alignment from the given phylogeny. 
To do so, we must specify the DNA root sequence and a mutation rate. 
In this example, the DNA root sequence consists in four blocks of 250 nucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.

\begin{lstlisting}[
    language=R,
    floatplacement=ht, frame=single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment.}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

By default, an alignment is created using a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model. 
A JC site model assumes that mutation rates from any nucleotide to any other 
are equal and constant. 
A strict clock model assumes that the mutation rates of all species are equal and constant.
Due to this, we state that the generative model for the alignment is
a JC site model and a strict clock model.

In the second step we state our experiments. In this context, we
define an experiment as a combination of an inference model
and the conditions under which a Bayesian inference is executed.
Within this example, we specify that our experiment uses the
generative model (which means that the inference model will be 
a combination of JC site model, strict clock model and Yule tree prior), 
it will always be run and we skip the evidence measurement for our model.

@@

Due to sensible defaults, specifying this
results in:
\giovanni{Would it be clearer not to use sensible defaults where we call the functions with arguments we refer to in the main text? For example here it could be clearer for the reader to see:
create\_experiment(site\_model = create\_jc69\_site\_model(), clock\_model = create\_strict\_clock\_model(), tree\_prior = create\_yule\_tree\_prior())
}

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create a default experiment, as described in Table~\ref{tbl:RQ1}.
  }
]
experiments <- list(create_experiment())
\end{lstlisting}

\verb;pirouette; has a plotting function used for convenience:

@@

The resulting figure is shown in figure~\ref{fig:example_1}.

@@

The error distribution obtained in this way can serve as a control:
imagine one develops a novel tree prior related to the Yule model.
One can predict that BEAST2 will be worse in recovering the phylogenies
from that novel tree prior, as only the existing tree priors are used
in inference. 
To determine how much worse the recovery is, one needs to know the 
inference error made by existing tree priors. 
As this control is important, it is an inherent part of \verb;pirouette;:
it is called twinning, which is shown in example research question 3.

>>>>

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2()
\end{lstlisting}
For the option to select a best candidate model,
\verb;pirouette; needs the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

@@

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
These functions are demonstrated when in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example, and have sensible defaults.

@@

@@

\verb;pirouette; quantifies the influence of a tree prior,
by measuring the difference between a given tree and 
phylogenies in a posterior. Due to stochasticity, all posterior trees
will differ from the given phylogeny, even when the generative tree
prior G and alignment model A are the same as 
those in the inference model.
Measuring this difference results allows us to know the baseline error
of the \verb;pirouette; pipeline.
Because only standard tree priors are used in inference, 
and G must be used such a tree prior, it follows that
G must also be a standard tree prior.

@@

"What is the inference error made on phylogenies
created by a standard diversification model?"

In this example we use the Yule (pure-birth) tree prior 
as the generative tree prior G,
because it is the simplest standard tree prior. 
Table~\ref{tab:options} shows an overview of the standard
tree priors.
We choose to use a small tree with six taxa to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but does tie in with the mutation rate 
used in simulating an alignment in the next step.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_yule.png}
  \caption{The Yule tree, as created by listing~\ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
  \giovanni{Remove the text in top-left corner.}
  \richel{Agreed, iff the text matches the code}
  \giovanni{I'd use a sexier topology, purely for esthetical reasons.}
  \richel{Agreed, will pick a sexy RNG seed after create\_yule\_tree is there}
\end{figure}

The first step in \verb;pirouette; is to simulate a DNA alignment from the given phylogeny. 
To do so, we must specify the DNA root sequence, DNA mutation rate,
site model (in this context: a nucleotide substitution model) and a clock model. 
In this example, the DNA root sequence consists in four blocks of 250 nucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
Because the crown age of the phylogeny is 10 time units, each nucleotide will
have one (possibly-silent) mutation on average.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A Jukes-Cantor (JC, \cite{jukes1969evolution}) site model,
assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

\begin{lstlisting}[
    language=R,
    floatplacement=ht, frame=single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment.}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1,
  site_model = create_jc69_site_model(),
  clock_model = create_strict_clock_model()
)
\end{lstlisting}

As the site and clock model used here are also the defaults, 
these function arguments can be safely omitted: we just show these
explicitly for clarity.

In the second step we state our experiments, which will be just one.
We define an experiment X as a combination of an inference model I
and conditions, C.
In this example we pick I to be the same inference model as the generative one,
which is Yule tree prior, JC site model and strict clock model.
We specify C that the experiment will always be run.

@@

Due to sensible defaults, specifying this
results in:

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments
  }
]
experiments <- list(
  create_experiment(
    inference_conditions = create_inference_conditions(
      model_type = "generative", 
      run_if = "always"
    ), 
   inference_model = create_inference_model(
     site_model = create_jc69_site_model(), 
     clock_model = create_strict_clock_model(), 
     tree_prior = create_yule_tree_prior()
    )
  )
)
\end{lstlisting}

Due to sensible defaults, the same can be achieved by:

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create the experiments to answer research question 1, 
    using the default arguments
  }
]
experiments <- list(create_experiment())
\end{lstlisting}

The objects we just created need to be bundled
in a \verb;pirouette; parameters structure,
together with error measurement parameters E and
twinning parameters T. We use the default error measurement
parameters, which uses a burn-in of 10\% and the nLTT statistic to
measure the difference between phylogenies. We will not use twinning
here yet.  
All these arguments are bundled
and checked by \verb;create_pir_params;:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = create_error_measure_params(),
  twinning_params = NA
)
\end{lstlisting}

Now we can use our Yule tree and \verb;pir_params; to measure 
the inference error made on phylogenies
created by a standard diversification model:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
errors <- pir_run(
  phylogeny = phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

One the hardware we used, it will take approximately 10 minutes
to obtain the error distribution returned by \verb;pir_run;.
This error distribution is a data frame in the non-tidy short form.
Instead of converting it to tidy data and plotting it,
\verb;pirouette; has a plotting function that does exactly that:

@@

@@

The error distribution obtained is shown as a violin plot
in figure~\ref{fig:example_1}.

This error distribution has little relevance on its own. 
It only gives a baseline error made by
the Bayesian inference. We use 'a', because we have used only one Yule
tree. To obtain the baseline error, one should simulate multiple Yule
trees and merge their error distributions.

For this error distribution to have more relevance it needs to be compared
to an error distribution derived from a phylogeny 
derived from a non-standard speciation process.
In that way, the error distribution we have now serves as a control,
as it is the baseline error caused by stochasticity.
In this example, that non-standard speciation process must 
be closest related to a Yule process for the comparison to make sense.
Would the two error distributions be similar, the non-standard speciation 
process is of little influence.
Because measuring the baseline error is so important,
we made it an inherent part of \verb;pirouette;. 
We call it twinning and we use it in example research question 3.
